When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-07-21 code/Submissionv1.Rmd
commit changes

# First Layer: Topical Analysis

@@

```{r table, fig.align="center", include=TRUE, echo=FALSE}
countbytype %>%
  kable(format="latex", caption="Publications by Subject") %>%
  kable_styling(latex_options = "striped", font_size=10, full_width = T) %>% 
  row_spec(0, bold=T)
```

Table 1 reveals that an overwhelming majority (289 of 334, or 86.5%) of the total number of opinion pieces written since 1979 in the major newspapers examined can be classified according to one of the following three categories: **War_Ops**, **Trp_Clt**, and **Str_Adv**. It is reasonable to conclude that these op-eds written are appropriate from the perspective of civil-military relations in that they address topics which American citizens can reasonably expect retired military officers to bring their expertise to bear upon. The remaining 45 opinion pieces (13.5% of the total) fall into one of the other three categories: **MilPSoc**, **Dom_P**, and **Fig_Nme** and require further examination.  They are not necessarily inappropriate, but they likely address topics that directly address or are closely related to issues that could be considered as partisan or at least beyond appropriate military expertise. This second layer of analysis will help expose whether there is any indication that professionalism as a norm among retired military officers appears to be weakening.

\singlespace
# The Second Layer - Failing to Uphold Professionalism?
\doublespace
Of the remaining observations, two are removed upon immediate inspection. One was written by an unknown author and is entitled, "Fists Raised at West Point," and was published in _The Washington Post_ on May 13, 2016. It is removed because the article is actually a compendium of two separate letters sent to the editor by two different authors, one arguing for and the other against female West Point graduates who raised their fists upon graduating, which ignited some controversy about soon-to-be Army lieutenants purportedly making a symbol associated with the "Black Lives Matter" movement [@noauthor_fists_2016]. Because the two separate letters taken together do not take a firm position on the subject, but rather show a difference in opinion about a specific event happening in 2016, I remove the observation. The second piece I remove was written by former Democratic Congressman John Murtha, who was also a retired Marine Corps officer. As a politician, the American public can and should expect him to write in the capacity of a politician, and therefore, to make partisan arguments. Not wanting to skew the remaining data, I remove his publication for the remaining analysis. Thus, the second level of analysis proceeds on a total of 43 observations, which are listed and printed in chronological publishing order, in Table 2.

@@

```{r table2, fig.align = "center", include=TRUE, echo=FALSE}
  rem_obs %>%
  kable(format="latex", caption="Remaining Observations", longtable=T) %>%
  kable_styling(font_size=10, latex_options="striped", repeat_header_continued=T) %>%
  column_spec(1, width="12em") %>% 
  column_spec(2, width="16em") %>% 
  column_spec(3:4, width="2.5em") %>% 
  column_spec(5:7, width="3em") %>% 
  row_spec(0, bold=T)
```

```{r code, include=FALSE, echo=FALSE}
# Create a Variable Called case-count that counts the number of times by publishing year an article published 
# regarding social policy or general parties or parties by name or administrations occurs
adj2dataset <- adjdataset %>% group_by(PubYr) %>% 
  mutate(
    case_count=sum(MilPSoc ==1 | Dom_P==1 | Fig_Nme==1), 
    yr_total_count=sum(War_Ops==1 | Trp_Clt==1| Str_Adv==1 | MilPSoc ==1 | Dom_P==1 | Fig_Nme==1),
    prop_yr=(case_count/yr_total_count)
  ) %>% view()
```
\doublespace

A couple of interesting observations are worth noting about the data in Table 2. First, one can quickly observe that there appear to be three distinct clusters of observations with respect to publishing year. The first occurs in the time frame between roughly 1998-1999, the second between 2004-2007, and the third between 2016-2019. The clusters can also be seen in time in Figure 2 on the following page.

@@

```{r fig2, fig.height=4, fig.width=6, fig.align="center", include=TRUE, echo=FALSE}
datatograph <- data.frame(
  year=c("1994","1996","1998","1999","2000","2003", "2004", "2006", "2007", "2010", "2013","2014", "2016", "2017", "2018", "2019", "2020"), count=c(1, 1,2, 3, 1, 1, 3, 3, 3, 1, 1, 1, 6, 5, 4, 6, 1)
)
ggplot(datatograph, aes(x =year, y =count)) + geom_bar(stat = "identity", width=.2) +
  labs(
    x = "Year",
    y = "Count",
    color = NULL,
    title = "Figure 2: # of Op-Eds Written By Military Retirees",
    subtitle = "On Topics of Social Policy, Domestic Policy, or Political Administrations/Figures"
  )
```

```{r practice, include = TRUE, echo=FALSE}

plot1<- ggplot(data=adj2dataset, mapping=aes(x=PubYr, y=case_count)) + geom_point() + geom_smooth() + labs(x="Year", y="Count", color=NULL, title="Figure 2: Yearly # of Op-Eds Published by Retired Military Officers", subtitle="On Topics of Social Policy, Domestic Policy, or Political Administrations/Figures")
plot2 <- ggplot(data=adj2dataset, mapping=aes(x=PubYr, y=prop_yr)) + geom_point() + geom_smooth() + labs(x="Year", y="Count", color=NULL, title="Figure 2: Proportion of Total Op-Eds Published by Retired Military Officers ", subtitle="On Topics of Social Policy, Domestic Policy, or Political Administrations/Figures")
grid.arrange(plot1, plot2, ncol=2)
```

@@

Using these characteristics as guidelines, I therefore develop two new variables. The first is **Result**, which can take on the following values:
\singlespace
* **Permissible** -if the reader cannot clearly draw conclusions regarding the author's political stances or partisan preferences; or if the topic addressed in the publication in fact does in fact fall underneath the area of clear military expertise.
* **Questionable** -if the reader can draw some general, but not final, conclusions regarding the author's political stances or partisan preferences; or if the topic addressed in the publication is questionably an area of reasonable military expertise.
* **Inappropriate** -if the reader can draw clear conclusions regarding the author's partisan preferences or political leanings; or if the author addresses a topic that is not an area considered to be directly relevant to military expertise, or a matter that the military should judge.

\doublespace
The second new variable I develop is **Justification** in which I briefly comment on why I assigned the value of **Result** for the observation. Here I simply comment on whether it is a concern of partisan preferences being revealed and/or topic addressed that are responsible for the value of **Result**. The results of this second level of analysis are presented in Table 3. Note that several publications are deemed to take on values of permissible under the variable **Result**. Importantly, three of these pieces occurred during the first previously identified cluster, 1998-1999. No pieces from the second cluster, 2004-2007 have been deemed permissible, whereas in the third cluster, 2016-2019, three take on the value of permissible. Thus, we now have a picture of how prevalent the publication is of opinion pieces authored by retired military officers in major newspapers which have failed to uphold the norm of non-partisanship and apolitical position.

@@

```{r table3, include=TRUE, echo=FALSE}
secondlayer_obs %>%
  kable(format="latex", caption="Concern for Op-Ed Pieces Authored By Retired Military Officers in US Newspapers, 1979-2020", longtable=T) %>%
  column_spec(1, width="9.5em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3:4, width="2.5em") %>%
  column_spec(5, width="5em") %>%
  column_spec(6, width="6em") %>%
  kable_styling(latex_options="striped", font_size=10) %>% 
  row_spec(0, bold=T)
```
\doublespace

Figure Two displays the number of opinion pieces authored by retired military officers which failed to uphold the strict standards of professionalism, as espoused by Retired Generals MacArthur and Dempsey, within each applicable year. The results reveal several conclusions. The first is that it clearly shows, in a broad sense, that from 1979 through the first half of 2020, opinion pieces penned at the hand of retired military officers which violate the concepts of professionalism in questionable or inappropriate ways are the exception, not the norm. Indeed, the entire count of questionable or inappropriate publications after conducting the second level of analysis drop to just 30 of 334 total pieces, or 8.9% of the entire sample. Thus, ample support exists to confirm the first hypothesis, that the vast majority of opinion pieces authored by retired military officers will be apolitical and non-partisan in nature.

```{r fig3, fig.height=4, fig.width=5, fig.align="center", include=TRUE, echo=FALSE}
datatograph <- data.frame(
  year=c("1998","2000","2003", "2004", "2006", "2007", "2016", "2017", "2018", "2019", "2020"), count=c(2, 1, 1, 3, 3, 1, 5, 4, 4, 5, 1)
)
ggplot(datatograph, aes(x =year, y =count)) + geom_bar(stat = "identity", width=.2) +
  labs(
    x = "Year",
    y = "Count",
    color = NULL,
    title = "Figure 3: # of Op-Eds Written By Military Retirees",
    subtitle = "Failed to Uphold Professional Standards"
  )
```

@@

The third hypothesis, that the number of published opinion pieces which have strained the concept of professionalism has increased, even as political polarization has worsened in the past several decades, is also confirmed, as evidenced by both by Table 3 and Figure 2. It is interesting to note that according to Figure 2, the first publishing year in which an opinion piece penned by a retired officer failed to uphold professional standards occurred in 1998, nearly twenty years into (or almost half) of the entire sample. Why is this the case? Retired military officers in the years preceding 1998 had plenty of reasons to go to the press and air disagreements with political leaders, or to address important topics that may have pushed the limits of professionalism as has been defined in this paper. In fact, many of these officers would have been veterans of the Vietnam and Korean Wars, and perhaps, like Maxwell Taylor, who is in our sample, veterans perhaps of World War Two. At this point, though it is only speculation, one can speculate that the factors of military credibility and political polarization, however imprecise and hazy at this point, may have played a role in keeping the majority of an earlier generation's retired officers out of the popular press.

@@

```{r fig4, fig.height=4, fig.width=5, fig.align="center", include=TRUE, echo=FALSE}
datatograph <- data.frame(
  decade=c("1979-1989", "1989-1999", "1999-2009", "2009-2019" ), numcount=c(13, 51, 102, 161)
)
ggplot(datatograph, aes(x =decade, y =numcount)) + geom_bar(stat = "identity", width=.2) +
  labs(
    x = "Period",
    y = "Published Opinion Pieces",
    color = NULL,
    title = "Figure 4: # of Op-Eds Written By Military Retirees",
    subtitle = "Published in WSJ, NY Times, Wa Post, USA Today, and LA Times"
  )
```

@@

The goal of this paper was to explore and examine the ways in which political polarization and institutional credibility have impacted the professionalism of the military over the past fifty years. Examining the published opinion pieces authored by retired military officers in major US newspapers revealed that together, institutional credibility and political polarization have generated changes in the professionalism of some retired military officers, and more broadly, in how they as elites are communicating with the American public. Though the data examined demonstrates that the vast majority of op-eds published by retired military officers in the past fifty years exhibited professional standards, the data also showed two worrisome trends: that retired military officers are arguing about topics that fall outside of the traditional umbrella of national security, and that they are expressing clear partisan preferences at a greater rate than they previously did. Moreover, both of these trends have increased under the past four years of Donald Trump's Presidency.

>>>>

# First Layer Analysis: Topic Addressed

@@

```{r table2, fig.align="center", include=TRUE, echo=FALSE}
countbytype %>%
  kable(format="latex", caption="Publications by Subject") %>%
  kable_styling(latex_options = "striped", font_size=10, full_width = T) %>% 
  row_spec(0, bold=T)
```

Table 2 reveals that an overwhelming majority (289 of 334, or 86.5%) of the total number of opinion pieces written since 1979 in the major newspapers examined can be classified according to one of the following three categories: **War_Ops**, **Trp_Clt**, and **Str_Adv**. It is reasonable to conclude that these op-eds written are appropriate from the perspective of civil-military relations in that they address topics which American citizens can reasonably expect retired military officers to bring their expertise to bear upon. The remaining 45 opinion pieces (13.5% of the total) fall into one of the other three categories: **MilPSoc**, **Dom_P**, and **Fig_Nme** and require further examination.  They are not necessarily inappropriate, but they likely address topics that may be beyond reasonable military expertise or which reveal partisan preferences. This second layer of analysis will help expose whether there is any indication that professionalism as a norm among retired military officers appears to be weakening.

\singlespace
# The Second Layer - Failing to Uphold Professionalism?
\doublespace

Of the remaining observations, two are removed upon immediate inspection. One was written by an unknown author and is entitled, "Fists Raised at West Point," and was published in _The Washington Post_ on May 13, 2016. It is removed because the article is actually a compendium of two separate letters sent to the editor by two different authors, one arguing for and the other against female West Point graduates who raised their fists upon graduating, which ignited some controversy about soon-to-be Army lieutenants purportedly making a symbol associated with the "Black Lives Matter" movement [@noauthor_fists_2016]. Because the two separate letters taken together do not take a firm position on the subject, but rather show a difference in opinion about a specific event happening in 2016, I remove the observation. The second piece I remove was written by former Democratic Congressman John Murtha, who was also a retired Marine Corps officer. As a politician, the American public can and should expect him to write in the capacity of a politician, and therefore, to make partisan arguments. Not wanting to skew the remaining data, I remove his publication for the remaining analysis. Thus, the second level of analysis proceeds on a total of 43 observations, which are listed and printed in chronological publishing order, in Table 2.

@@

```{r table3, fig.align = "center", include=TRUE, echo=FALSE}
  rem_obs %>%
  kable(format="latex", caption="Remaining Observations", longtable=T) %>%
  kable_styling(font_size=10, latex_options="striped", repeat_header_continued=T) %>%
  column_spec(1, width="12em") %>% 
  column_spec(2, width="16em") %>% 
  column_spec(3:4, width="2.5em") %>% 
  column_spec(5:7, width="3em") %>% 
  row_spec(0, bold=T)
```

```{r code, include=FALSE, echo=FALSE}
# Create a Variable Called case-count that counts the number of times by publishing year an article published 
# regarding social policy, domestic policy, or political figures/administrations occurs
adj2dataset <- adjdataset %>% group_by(PubYr) %>% 
  mutate(
    case_count=sum(MilPSoc ==1 | Dom_P==1 | Fig_Nme==1), 
    yr_total_count=sum(War_Ops==1 | Trp_Clt==1| Str_Adv==1 | MilPSoc ==1 | Dom_P==1 | Fig_Nme==1),
    prop_yr=(case_count/yr_total_count)
  ) %>% view()
```
\doublespace

A couple of interesting observations are worth noting about the data in Table 3. First, one can quickly observe that there appear to be three distinct clusters of observations with respect to publishing year. The first occurs in the time frame between roughly 1998-1999, the second between 2004-2007, and the third between 2016-2019.

@@

@@

Using these characteristics as guidelines, I therefore develop two new variables. The first is **Result**, which can take on one of three values: the following values: **Permissible,** if the reader cannot clearly draw conclusions regarding the author's political leaning or partisan preferences; or if the topic addressed in the publication, after a second reading, clearly falls within military expertise; **Questionable,** if the reader can draw some but not complete conclusions regarding the author's political leaning or partisan preferences, or if the topic addressed in the publication is questionably an area of reasonable military expertise; and **Inappropriate,** if the reader can draw clear conclusions regarding the author's partisan preferences or political leanings, or if the author addresses a topic that clearly falls outside of traditional military expertise. The second new variable I develop is **Justification.** In this variable, a brief explanation is placed which addresses why an observation is assigned a particular value of **Result.**

The results of this second level of analysis are presented in Table 3. Note that several publications are deemed to take on values of permissible under the variable **Result**. Importantly, three of these pieces occurred during the first previously identified cluster, 1998-1999. No pieces from the second cluster, 2004-2007 have been deemed permissible, whereas in the third cluster, 2016-2019, three take on the value of permissible. Thus, we now have a picture of how prevalent the publication is of opinion pieces authored by retired military officers in major newspapers which have failed to uphold the norm of non-partisanship and apolitical position.

@@

```{r table4, include=TRUE, echo=FALSE}
secondlayer_obs %>%
  kable(format="latex", caption="Op-Ed Pieces By Retired Military Officers which failed to Uphold Professional Standards, 1979-2020", longtable=T) %>%
  column_spec(1, width="9.5em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3:4, width="2.5em") %>%
  column_spec(5, width="5em") %>%
  column_spec(6, width="6em") %>%
  kable_styling(latex_options="striped", font_size=10) %>% 
  row_spec(0, bold=T)
```
\doublespace

Figure 2 displays the number of opinion pieces authored by retired military officers which failed to uphold the strict standards of professionalism, as espoused by Retired Generals MacArthur and Dempsey, within each applicable year. The results reveal several conclusions. The first is that it clearly shows, in a broad sense, that from 1979 through the first half of 2020, opinion pieces penned at the hand of retired military officers which violate the concepts of professionalism in questionable or inappropriate ways are the exception, not the norm. Indeed, the entire count of questionable or inappropriate publications after conducting the second level of analysis drop to just 30 of 334 total pieces, or 8.9% of the entire sample. Thus, ample support exists to confirm the first hypothesis, that the vast majority of opinion pieces authored by retired military officers will be apolitical and non-partisan in nature.

```{r fig2, fig.height=4, fig.width=5, fig.align="center", include=TRUE, echo=FALSE}
datatograph <- data.frame(
  year=c("1998","2000","2003", "2004", "2006", "2007", "2016", "2017", "2018", "2019", "2020"), count=c(2, 1, 1, 3, 3, 1, 5, 4, 4, 5, 1)
)
ggplot(datatograph, aes(x =year, y =count)) + geom_bar(stat = "identity", width=.2) +
  labs(
    x = "Year",
    y = "Count",
    color = NULL,
    title = "Figure 2: # of Op-Eds Written By Military Retirees",
    subtitle = "Failed to Uphold Professional Standards"
  )
```

@@

The third hypothesis, that the number of published opinion pieces which have strained the concept of professionalism has increased, even as political polarization has worsened in the past several decades, is also confirmed, as evidenced by both by Table 3 and Figure 3. It is interesting to note that according to Figure 3, the first publishing year in which an opinion piece penned by a retired officer failed to uphold professional standards occurred in 1998, nearly twenty years into (or almost half) of the entire sample. But why is this the case? Retired military officers in the years preceding 1998 had plenty of reasons to go to the press and air disagreements with political leaders, or to address important topics that may have pushed the limits of professionalism as has been defined in this paper. In fact, many of these officers would have been veterans of the Vietnam and Korean Wars, and perhaps, like Maxwell Taylor, who is in our sample, veterans perhaps of World War Two. At this point, though it is only speculation, one can speculate that the factors of military credibility and political polarization, however imprecise and hazy at this point, may have played a role in keeping the majority of an earlier generation's retired officers out of the popular press.

@@

```{r fig3, fig.height=4, fig.width=5, fig.align="center", include=TRUE, echo=FALSE}
datatograph <- data.frame(
  decade=c("1979-1989", "1989-1999", "1999-2009", "2009-2019" ), numcount=c(13, 51, 102, 161)
)
ggplot(datatograph, aes(x =decade, y =numcount)) + geom_bar(stat = "identity", width=.2) +
  labs(
    x = "Period",
    y = "Published Opinion Pieces",
    color = NULL,
    title = "Figure 3: # of Op-Eds Written By Military Retirees",
    subtitle = "Published in WSJ, NY Times, Wa Post, USA Today, and LA Times"
  )
```

@@

The goal of this paper was to explore and examine the ways in which political polarization and institutional credibility have impacted the professionalism of the military over the past fifty years. Examining the published opinion pieces authored by retired military officers in major US newspapers revealed that together, institutional credibility and political polarization have generated changes in the professionalism of some retired military officers, and more broadly, in how military elites communicate with the American public. Though the data examined demonstrates that the vast majority of op-eds published by retired military officers in the past fifty years exhibited professional standards, the data also showed two worrisome trends: that retired military officers are arguing about topics that fall outside of the traditional umbrella of national security, and that they are expressing clear partisan preferences at a greater rate than they previously did. Moreover, both of these trends have increased under the past four years of Donald Trump's Presidency. These trends are captured in Figure 4, below, which shows the number (left) and proportion (right) of opinion pieces authored by retired military officers over the past 40 years that addressed issues of social policy, domestic policy, or specific political figures or administrations. The graphs tell an interesting story: the number and proportion were both basically 0 40 years ago, then increased slightly until about 2003, at which point, both decreased until roughly 2011. After that, both the number and proportion have risen sharply. Nearly 40% of all opinion pieces in 2019 published in major US newspapers by retired military officers were written about topics that fell outside of traditional national security topics.

```{r build plots, include = FALSE, echo=FALSE}
plot1<- ggplot(data=adj2dataset, mapping=aes(x=PubYr, y=case_count)) + geom_point() + geom_smooth() + labs(x="Year", y="Number", color=NULL)
plot2 <- ggplot(data=adj2dataset, mapping=aes(x=PubYr, y=prop_yr)) + geom_point() + geom_smooth() + labs(x="Year", y="Proportion", color=NULL)
```

```{r figure 4, include=TRUE, warning= FALSE, message = FALSE, echo=FALSE}
grid.arrange(plot1, plot2, ncol=2, top="Figure 4: #/Proportion of Op-Eds Published by Retired Military Officers, 1979-2020", bottom="On Topics of Social Policy, Domestic Policy, and Political Administrations/Figures")
```
